# How matching works

This chapter explains gnafr from the inside: how a raw string becomes structured
components, how those components become a score, and how `gnaf_match()` decides
which database rows are even worth scoring. Every number quoted here was read from
the current source ([`R/parse.R`](../R/parse.R), [`R/score.R`](../R/score.R),
[`R/score_metrics.R`](../R/score_metrics.R), [`R/gnaf_match.R`](../R/gnaf_match.R))
and, where practical, verified by actually running it —
look for the executed code blocks.

```{mermaid}
%%{init: {'themeVariables': {
  'fontSize': '12px',
  'nodePadding': 6,
  'rankSpacing': 20,
  'edgeSpacing': 8,
  'diagramPadding': 6
}}}%%
flowchart LR

    A[gnaf_match input]
    B{Exact label<br/>match?}
    C{In match<br/>cache?}
    D[address_parse]

    A --> B
    B -- yes --> Z[Result]
    B -- no --> C
    C -- yes --> Z
    C -- no --> D

    D --> E{Postcode<br/>parsed?}

    subgraph P[DuckDB Matching]
        direction TB
        F[Postcode path]
        G[State path]
    end

    E -- yes --> F
    E -- no --> G

    F --> H{Best score <=<br/>fallback_threshold<br/>or unmatched?}
    G --> H

    H -- no --> Z
    H -- yes & locality parsed --> I[Locality fallback<br/>fuzzy suburb → postcode]

    I --> H2{Best score <=<br/>fallback_threshold<br/>or unmatched?}
    H2 -- no --> J
    H2 -- yes --> I2[Street-number-relaxed fallback<br/>drop number filter, search by street name]
    I2 --> J

    J{Still unmatched &<br/>street_only_fallback=TRUE?}

    J -- no --> Z
    J -- yes --> K[Street-only fallback]

    K --> Z
```

## Parsing {#sec-parsing}

`address_parse()` turns a raw string into a row of structured fields:
`in_postcode`, `in_state`, `in_locality`, `in_street_name`, `in_street_type`,
`in_street_suffix`, `in_number_first`, `in_number_last`, `in_number_suffix`,
`in_flat_type`, `in_flat_number`, `in_building_name`.

It runs in two tiers:

1. **Vectorised fast paths** (`.parse_vectorized`) — four `stringi`-based regex
   patterns, applied across the *whole* input vector at once, that together cover
   roughly 95%+ of well-formed Australian addresses:
   - **4a — slash notation**: `FLAT/NUMBER STREETNAME` (`"110/120 Musgrave Rd"`)
   - **4b — flat-type prefix**: `UNIT NUMBER NUMBER STREETNAME` (`"Unit 3 120 Musgrave Rd"`)
   - **4b2 — attached single-letter prefix**: `F8 536 STREETNAME` (`F`=Flat, `A`=Apartment, `U`=Unit; any other letter defaults to Unit)
   - **4c — simple form**: `NUMBER[-NUMBER] STREETNAME` (the most common case by far), with an "implied pair" sub-case for the common Australian convention `<unit> <number> <street>` written with no explicit `UNIT` keyword (`"10 120 Musgrave Rd"`)
2. **A row-by-row fallback** (`.parse_single`) for anything the fast patterns don't
   recognise — uncommon structures, fuzzy-misspelt street types, addresses with no
   identifiable street type at all.

Before any of that, every input has its trailing **postcode/state** stripped (in
either order — `"...QLD 4059"` or `"...4059 QLD"`), then its rightmost **street
type** token located.

::: {.callout-tip}
## The comma hint
If the original string has a comma (`"123 St James Rode, Tamborine Mountain..."`),
the word immediately before the *last* comma is treated as a strong hint for the
street type — even fuzzily, via Jaro-Winkler against the known type list. This
exists specifically to catch two failure modes of a plain rightmost-match search:
coincidental abbreviation collisions inside a multi-word street name (`"St"` inside
`"St James Rode"` matching the `STREET` abbreviation and hiding the real, misspelt
type `"Rode"`), and misspelt types that a pure rightmost scan would miss entirely.
:::

### Live example

```{r}
library(gnafr)

address_parse(c(
  "unit 110 120 musgrave Road red hill 4000 QLD",
  "U110 1120 musgrave rd red hill 4000",
  "Cambridge on the hill 110/120 musgrave road red hill QLD 4000",
  "18-20 drift cl goldsborough QLD 4865",
  "F8 536 example parade brisbane 4000"
))
```

### Normalisation (`normalize = TRUE`, the default)

After structural parsing, common abbreviations in `in_locality` and `in_street_name`
are expanded to the form G-NAF actually uses, so they compare equal at scoring time
instead of merely *similar*:

| Pattern | Expands to |
|---|---|
| `MNT`, `MT` | `MOUNT` |
| `ST` (as a whole word, after the street type has already been extracted) | `SAINT` |
| `NTH`, `STH`, and a leading `N`/`S`/`E`/`W` | `NORTH` / `SOUTH` / `EAST` / `WEST` |
| `CK` | `CREEK` |
| `&` | `AND` |
| Ordinal numerals (`1ST`…`20TH`) | written words (`FIRST`…`TWENTIETH`), street name only |

```{r}
address_parse("Mt Gravatt East", normalize = TRUE)[, .(in_locality)]
address_parse("Mt Gravatt East", normalize = FALSE)[, .(in_locality)]
```

This matters because the suburb/street-name score is Jaro-Winkler *similarity*, not
identity — `"MOUNT GRAVATT EAST"` vs `"MT GRAVATT EAST"` would still score well, but
expanding it first means a correctly-typed-but-abbreviated suburb scores a perfect
match instead of a near-miss.

## Scoring {#sec-scoring}

Every (input, candidate) pair is scored on six components that sum to `total_score`
(0–100). The same logic exists twice in the source — once as DuckDB SQL `CASE`
expressions (`.score_sql_exprs()`, used by every matching path so scoring happens
inside the database), and once as equivalent `data.table` vector operations
(`.score_pairs()`, used for the exact-label fast path). **The current default
weights** (`R/score.R`):

```{r}
#| echo: false
w <- list(postcode = 20, suburb = 15, street_name = 40, street_type = 10, number = 10, flat = 5)
data.frame(Component = names(w), Weight = unlist(w)) |> knitr::kable()
```


You can pass your own `weights` list to bias matching — e.g. toward street number for
a number-sensitive workflow — but note this **disables the match cache for that
call** (cached scores were computed under the default weights, so reusing them under
different weights would silently misreport).

### Postcode (default weight 20)

Exact match scores full weight. A difference of 1, 2, or 3 retains the existing
70%, 40%, 20% partial-credit tiers. Outside those tiers, a single adjacent digit
transposition receives 40% when the leading digit is unchanged: `4067` versus
`4076` scores 8/20. Other digit substitutions and swaps involving the leading
digit receive zero. Postcodes are compared as four digits, preserving the leading
zero in Northern Territory codes. Missing or invalid values receive zero.

In practice this only ever fires on the **locality fallback** path (@sec-matching-paths):
the primary postcode-path query joins `ON g.postcode = i.in_postcode`, so by
construction every row it scores already has `score_postcode` at full weight. The
partial-credit branches only become reachable once the locality fallback widens the
join to nearby postcodes (±3) or to whatever postcode a fuzzy-matched locality name
points at, regardless of how far off the stated postcode was:

```{r}
con <- gnaf_connect(":memory:")
gnaf_init(con)
gnaf_add(con, data.table::data.table(
  address_label = "10 MUSGRAVE ROAD, RED HILL QLD 4059",
  number_first = 10L, street_name = "MUSGRAVE", street_type = "ROAD",
  locality_name = "RED HILL", state = "QLD", postcode = 4059L
))

# Postcode typo'd by one digit (4058 vs 4059) — the number and street still
# match exactly, so the locality fallback finds it and scores the postcode
# component at 70% of weight (round(20 * 0.7) = 14) instead of rejecting it.
gnaf_match("10 Musgrave Road, Red Hill QLD 4058", con, verbose = FALSE)[
  , .(matched, total_score, score_postcode, address_label)
]
```

### Suburb / locality (default weight 15) and street name (default weight 40)

Both combine two metrics with equal weight:

1. **Jaro–Winkler**, reshaped to suppress its unrelated-name noise floor:
   `J = 0.05 + 0.95 * clamp((JW - 0.6) / 0.4, 0, 1)^2`.
2. **Damerau–Levenshtein**, which counts insertions, deletions, substitutions
   and adjacent transpositions: `E = 1 - distance / max(lengths)`.

The credit fraction is `max(0.05, 0.5 * J + 0.5 * E^2)`. This balances common
prefixes against edits across the whole name. For example, input `XILLIAM`
scores 27/40 against `WILLIAM`, ahead of 25/40 against `XILLIAMSON`. Previously,
the prefix-heavy candidate won 29/40 versus 24/40.

Opposing whole-word directions (`NORTH`/`SOUTH` or `EAST`/`WEST`) halve the
fraction. This applies independently to each name bucket: `MOUNT GRAVATT EAST`
versus `MOUNT GRAVATT WEST` scores 17/40 as a street name, previously 36/40.
Words such as `EASTON` are not directions. Names containing both opposites on
an axis do not assert a single direction on that axis.

Case and whitespace are normalised before comparison. Full credit requires
equal, non-empty normalised text. Imperfect names are capped at one point below
the rounded component weight (with a floor of zero), so integer rounding cannot
turn a small difference into exact agreement. Missing or blank names receive zero.

DuckDB uses its built-in metrics. R uses `fast.string::jaro_winkler()` and
`stringdist(..., method = "dl", useBytes = TRUE)` on UTF-8 text, matching DuckDB's
byte-based comparisons and length denominator. The candidate-pruning bound uses
the same final name score, so it cannot discard a candidate whose edit evidence
raises it above `min_score`.

Metric references: [DuckDB text functions](https://duckdb.org/docs/lts/sql/functions/text)
and the [stringdist manual](https://markvanderloo.r-universe.dev/stringdist/doc/manual.html).

For example, `24 ILLAWONG` must not receive the same street-name credit as
`ILLAWONG`. The former can arise when a unit marker is misspelled and a unit
number is mistaken for a street number. A unit marker is not required:
`MY BUILDING NAME 3 24 ILLAWONG STREET` is parsed as building name
`MY BUILDING NAME`, unit `3`, street number `24`, and street `ILLAWONG`.
The vectorised and scalar parsers share this trailing-number-pair rule.
Street-number candidate filtering therefore uses `24`, not the unit number.
The parser also recovers unambiguous
one-edit or adjacent-transposition variants of `UNIT`, `FLAT`, `APARTMENT`,
and `SUITE` when followed by a unit number, a street number, and a street name.
It preserves the original text in `input_raw`. A corrected marker can still
produce a score of 100 when the standardised components agree exactly.

Scores are weighted agreement measures, not calibrated probabilities or a
guarantee of a unique address. The equal metric blend is a heuristic verified
against regression examples, not fitted to labelled production matches. It raises
some historical scores and lowers others; assess thresholds against labelled
examples for your population. Versioned cache entries from previous scoring rules
are ignored. `dev/benchmark_scoring.R` compares examples and scoring runtime with
a Git revision without requiring a production database.

### Street type (default weight 10)

A three-tier comparison after mapping both sides through the street-type
dictionary. `RD`/`ROAD` and `ST`/`STREET` receive full credit even when a stored
candidate still contains an abbreviation. Distinct known types keep their
existing disagreement tier:

- Both sides present and equal → full weight
- Either side missing a type (including both) → 50% of weight
- Both sides have a type, but they disagree → 40% of weight

That 40%-for-disagreement floor is deliberate: a wrong street type (`"Avenue"`
typed where G-NAF has `"Drive"`) is a very common, low-stakes user error and
shouldn't fully cancel out an otherwise strong match.

```{r}
gnaf_add(con, data.table::data.table(
  address_label = "13-27 FAIRWAY DRIVE, CLEAR ISLAND WATERS QLD 4226",
  number_first = 13L, number_last = 27L, street_name = "FAIRWAY", street_type = "DRIVE",
  locality_name = "CLEAR ISLAND WATERS", state = "QLD", postcode = 4226L
))

gnaf_match("13 Fairway Avenue, Clear Island Waters QLD 4226", con, min_score = 0, verbose = FALSE)[
  , .(matched, total_score, score_street_type, address_label)
]
```

Street directions also qualify this component: matching directions retain the
street-type credit, a direction missing on one side halves it, and conflicting
directions score zero. Codes such as `N` and `NTH` compare equal to `NORTH`.

### Street number (default weight 10)

Both input and candidate ranges are treated as inclusive intervals:

| Relationship | Example input → candidate | Fraction of number weight |
|---|---|---:|
| Exact number or interval | `10–20` → `10–20` | 100% |
| Candidate contains input | `12` → `10–20` | 70% |
| Input contains candidate | `10–20` → `12` | 50% |
| Partial overlap | `10–20` → `18–25` | 30% |
| Disjoint intervals or missing input number | `10–20` → `30` | 0% |

An explicit street number takes precedence over a lot in both retrieval and
number scoring. A lot identifier is used only when the input has no street
number, and requires agreement after removing numeric zero padding (`007`
equals `7`, and `007A` equals `7A`). Alphabetic suffixes and compound identifiers
remain significant; `7A` does not equal `7B`. Candidate lot filtering uses the
same normalisation as scoring. A `LOT 7` label cannot supply a fallback street
number of `7`. `gnaf_match_features()` checks lot conflicts independently even
when a street number is present; a high agreement score is not a probability.
The first point in a larger candidate range receives containment credit, rather
than an exact score. Ranges do not imply odd/even parity, and matching returns
existing records without creating synthetic addresses.

A supplied number suffix must agree with the house-number token immediately
before the candidate's street name. This works with prefixes such as
`UNIT 2 LEVEL 3 10A MAIN ROAD`; a unit numbered `10A` cannot substitute for house
`10B`. An omitted input suffix halves number credit when the candidate supplies
one. The database does not store house-number suffixes separately, so suffix
matching still depends on a usable address label. The parser currently retains
only the first endpoint's suffix in a range.

```{r}
gnaf_match("15 Fairway Drive, Clear Island Waters QLD 4226", con, verbose = FALSE)[
  , .(matched, total_score, score_number, address_label)
]
```

::: {.callout-warning}
## The number filter is a gate, not a soft score
Unlike every other component, an unmatched street number doesn't just *score* zero —
in the postcode, state and locality-fallback paths, candidates whose number doesn't
satisfy the pre-filter (intersecting numeric intervals, or a suffixed house number
recovered from the candidate label) are excluded from the SQL join entirely and
never reach scoring at all. Disabling every fallback below shows the gate in its raw
form: an input with a house number that simply isn't in G-NAF for that street comes
back `matched = FALSE`, `match_status = "no_candidate"` — not a low-scoring row you
can raise by lowering `min_score`:

```{r}
gnaf_match("999 Musgrave Road, Red Hill QLD 4059", local({
  c2 <- gnaf_connect(":memory:"); gnaf_init(c2)
  gnaf_add(c2, data.table::data.table(
    address_label = "10 MUSGRAVE ROAD, RED HILL QLD 4059",
    number_first = 10L, street_name = "MUSGRAVE", street_type = "ROAD",
    locality_name = "RED HILL", state = "QLD", postcode = 4059L
  ))
  c2
}), min_score = 0, street_number_fallback = FALSE, verbose = FALSE)[, .(matched, match_status, total_score)]
```

Two fallbacks exist precisely for this situation, on by default so it's rarely seen
in practice: the **street-number-relaxed fallback** (@sec-matching-paths) drops the
number constraint and re-searches by street name whenever the best result so far is
weak, letting the correct street compete on its own (honestly number-scored) merits —
the same input above resolves under the defaults used everywhere else in this
document. Because number and unit credit is scaled by street agreement
(@sec-street-gate), a different street that merely shares the input's number cannot
outrank it:

```{r}
gnaf_match("999 Musgrave Road, Red Hill QLD 4059", local({
  c2 <- gnaf_connect(":memory:"); gnaf_init(c2)
  gnaf_add(c2, data.table::data.table(
    address_label = "10 MUSGRAVE ROAD, RED HILL QLD 4059",
    number_first = 10L, street_name = "MUSGRAVE", street_type = "ROAD",
    locality_name = "RED HILL", state = "QLD", postcode = 4059L
  ))
  c2
}), min_score = 0, verbose = FALSE)[, .(matched, match_status, total_score, score_number, address_label)]
```

The **street-only fallback** (opt-in, `street_only_fallback = TRUE`) covers what the
street-number-relaxed fallback can't: it fires only once every number-aware path has
come up completely empty, and matches against pre-built number-free alias records, so
it can additionally recover a *misspelt* street name via those records' own fuzzy
resolution - useful when the street-number-relaxed fallback's exact street-name join
doesn't find anything either.
:::

### Flat / unit and level (default weight 5)

Flat and level identifiers are compared independently. Each earns full credit
for agreement, half for a missing identifier on one side, and zero for conflict.
When both dimensions occur on either side, 60% of the flat weight goes to units
and 40% to levels. Otherwise the present dimension receives the whole weight.
With both dimensions absent, the existing full credit is retained.

For matching identifiers, `UNIT`, `APARTMENT` and `FLAT` are equivalent, as are
`LEVEL` and `FLOOR`. Other conflicting supplied types halve that dimension's
credit. Case, surrounding whitespace and numeric zero padding are ignored:
`003` equals `3`, and `003A` equals `3A`. Different digits, alphabetic suffixes
and compound identifiers remain distinct. The combined component is rounded once, using the same rules
in R and DuckDB.

For input `UNIT 2 LEVEL 3`, unit 2 on level 3 scores 5, unit 2 with a missing level
scores 4, and unit 2 on level 4 scores 3. These are heuristic agreement scores,
not calibrated probabilities: even a high total can contain a unit conflict.
Use component scores and labelled examples to set acceptance criteria.

### Number and unit credit is conditional on the street {#sec-street-gate}

A house number or unit only locates an address *within* a street. Agreement on
either says nothing when the street itself is different, so `score_number` and
`score_flat` are scaled by a **street gate**: 1 for the same street, falling with
the same name similarity that drives `score_street_name`, and left at 1 when either
street name is missing (there is nothing to gate on, so lot-only inputs are
unaffected). A near-miss spelling keeps most of its number and unit credit; an
unrelated street keeps almost none.

Without the gate the components simply add up, so a *different* street that happens
to share the input's number and unit could collect the whole number-plus-flat weight
and outrank the correct street whose number isn't in G-NAF yet (for example because
new development lags the release). With it, that wrong street falls far below the
right one, which the street-number-relaxed fallback below retrieves and returns with
an honestly low `score_number`. When several addresses on the right street tie, the
one whose house number is nearest the input's ranks first, then the lowest PID.

The gate is part of the ranking score only. `total_score` is still the sum of its
component columns, and [`gnaf_match_features()`](05-function-reference.qmd) still
reports raw agreement for each component. Two spellings of one street can differ
slightly between R and DuckDB's Jaro-Winkler, so the R twin of the scorer (used for
exact-label matches and in tests) can differ by a point on a mis-spelt street; the
DuckDB scores are the ones `gnaf_match()` ranks by. Cache algorithm version 19
ignores scores computed without the gate.

The six public weights and score columns remain unchanged. Locality and alias
fallback cutoffs scale with custom weights, and cache algorithm version 12 ignores
scores computed by older versions.

## Matching paths {#sec-matching-paths}

`gnaf_match()` tries, in order, stopping as soon as an input is satisfied:

### 1. Exact label match

If the standardised input string equals a G-NAF `address_label` exactly, it is
scored. Further candidate search is skipped only when enough score-100 results
fill `max_results` — no parsing-driven join needed. Built for
re-processing addresses gnafr (or G-NAF itself) already standardised. Implemented as
a hash join against a registered temporary table rather than a giant SQL `IN (...)`
list, which keeps it fast even for tens of thousands of inputs.

### 2. Match cache

`gnaf_match_cache` stores `(input_standardised → address_detail_pid, scores)` for
every result that previously scored at or above `cache_threshold` (default 95).
Lookups join on `input_standardised` — the same building/flat/number/street/locality
string `gnaf_match()` itself derives from the parsed input — so a future call with a
*differently worded but equivalently parsed* address can still hit the cache. The
cache is **only consulted/written when `weights` are exactly the package defaults**;
non-default weights bypass it entirely for that call (see [Scoring](#sec-scoring)).

### 3. Postcode path

The hot path for the overwhelming majority of inputs: an equi-join on
`g.postcode = i.in_postcode` against the full address table, restricted by the
number pre-filter and a score upper bound before full component scoring. The
rounded street score plus the maximum possible contribution from the other
components must reach `min_score`. There is no fixed street-similarity floor:
zero street weight really means street agreement contributes nothing. This remains
deliberately a cheap, exact equi-join: near-miss postcodes are not retrieved here at
all (see the callout above) — that's the locality fallback's job.

### 4. State path

For inputs with no parseable postcode at all: the same query shape, joined on
`g.state = i.in_state` instead. Necessarily slower and noisier (a state-wide scan
instead of a postcode-blocked one), so it's only used when there's truly no postcode
to block on.

### 5. Locality fallback

For postcode-bearing inputs, this runs when no candidate survives or the best
postcode candidate is at or below `fallback_threshold` **and** has a weak locality
component (less than 85% of its weight, rounded). It also handles inputs with a
locality but neither postcode nor state. An already completed state search is not
repeated through locality fallback.

Exact locality names discover postcodes first. For noisy names, similarities of
at least 0.85 qualify; the five best **distinct postcodes** are retained, ranked by
their best locality similarity. Several aliases for one postcode occupy one slot.
Explicit state limits the locality index search. Postcode offsets of ?1, ?2 and
?3 are used only when no locality-derived postcode was found, and the original
postcode is excluded because it was already searched.

```{r}
gnaf_add(con, data.table::data.table(
  address_label = "77 BROADWATER ROAD, MOUNT GRAVATT EAST QLD 4122",
  number_first = 77L, street_name = "BROADWATER", street_type = "ROAD",
  locality_name = "MOUNT GRAVATT EAST", state = "QLD", postcode = 4122L
))

# Postcode is wildly wrong (9999); the suburb name (after MT -> MOUNT
# normalisation) is enough for the locality fallback to find it.
gnaf_match("77 Broadwater Road, Mt Gravatt East QLD 9999", con, min_score = 0, verbose = FALSE)[
  , .(matched, match_status, total_score, score_postcode, score_suburb, address_label)
]
```

### 6. Street-number-relaxed fallback (default: `street_number_fallback = TRUE`)

Every path above blocks candidates on the parsed street number (@sec-scoring's "gate,
not a soft score" callout) *before street name is ever scored*. So a real G-NAF row
on the exact right street, at a number just not present in the loaded data, is
invisible to paths 3-5 no matter how strong the rest of the match would be — the
algorithm can only ever surface whatever *other* row happened to satisfy the number
filter, including a coincidentally similar street name elsewhere.

This path runs when the best result across paths 3-5 is still at or below
`fallback_threshold`, or absent entirely. It drops the number constraint and joins on
`street_name` instead (postcode-anchored when the input has one, state-anchored
otherwise), so a real entry on the correct street becomes a scoreable candidate again
- with an honestly low `score_number` - and the existing scoring formula decides
whether it beats whatever else was found. Unlike street-only fallback below, the
street-name join here is exact, not fuzzy: it targets "right street, wrong number,"
not "misspelt street."

```{r}
local({
  c3 <- gnaf_connect(":memory:"); gnaf_init(c3)
  gnaf_add(c3, data.table::data.table(
    address_detail_pid = c("WILLIAM_REAL", "HICKEY_DECOY"),
    address_label = c("23 WILLIAM STREET, PORTSMITH QLD 4870",
                      "61 WILLIAM HICKEY STREET, REDLYNCH QLD 4870"),
    number_first = c(23L, 61L),
    street_name = c("WILLIAM", "WILLIAM HICKEY"), street_type = "STREET",
    locality_name = c("PORTSMITH", "REDLYNCH"), state = "QLD", postcode = 4870L
  ))
  # 61 William Street doesn't exist in this data - only 23 does. Without this
  # fallback, the number gate would instead surface "61 William Hickey
  # Street" (a different street/suburb that merely shares one word and has a
  # coincidentally matching number).
  out <- gnaf_match("61 William Street, Portsmith QLD 4870", c3, verbose = FALSE)[
    , .(address_detail_pid, total_score, score_number, score_street_name, address_label)
  ]
  gnaf_disconnect(c3)
  out
})
```

### 7. Street-only fallback (opt-in: `street_only_fallback = TRUE`)

For inputs still unmatched after every other path. Matches only against
`alias_type = 'street_only'` rows — synthetic, number-free address records derived
from G-NAF by [`gnaf_build_street_aliases()`](05-function-reference.qmd#gnaf_build_street_aliases)
(or the official PSV alias tables via
[`gnaf_load_psv()`](05-function-reference.qmd#gnaf_load_psv)). Its join filters only
on postcode and `alias_type`, so street name is scored (fuzzily, same as every
number-aware path) rather than filtered — meaning it can additionally recover a
*misspelt* street name that the street-number-relaxed fallback's exact join would
miss. Typical cause: a brand new subdivision whose house numbers G-NAF hasn't
published yet, or an input that genuinely has no street number.

```{r}
gnaf_disconnect(con)
```

## Why everything runs inside DuckDB

Every matching path registers the parsed inputs as a temporary DuckDB table
(`duckdb::duckdb_register()`), builds one SQL query that joins, scores, ranks
(`ROW_NUMBER() OVER (PARTITION BY input_id ...)`), and filters by `min_score` —
*then* pulls only the surviving rows into R. A few design choices fall directly out
of this:

- **No window-function aggregate before the score filter.** Computing something
  like `COUNT(*) OVER (...)` before filtering on `total_score` would force DuckDB to
  fully materialise the join first, which is exactly the memory blow-up bulk matching
  is built to avoid.
- **A conservative score upper bound precedes full scoring.** Pruning respects
  `weights` and `min_score`, including fractional and zero weights. Number and
  geographic blocking still restrict candidate retrieval.
- **Query failures are errors.** A missing table or failed SQL query must not look
  like an ordinary unmatched address just because `verbose = FALSE`.
- **Registered temp tables, not literal `IN (...)` lists**, for the exact-label and
  cache lookups — `duckdb_register()` lets DuckDB hash-join against an R
  `data.table` directly, which scales far better than a 50,000-item SQL literal.

## The database schema

`gnaf_init()` creates two structurally identical tables — `gnaf_addresses` and
`custom_addresses` — every G-NAF Core column plus the alias/source bookkeeping
columns gnafr adds (`source`, `alias_type`, `alias_principal`, `principal_pid`,
`primary_secondary`, `primary_pid`, `geocode_type`), three supporting structures, and
runs any schema migrations needed for databases created by an older version of
gnafr (every `ALTER TABLE ... ADD COLUMN IF NOT EXISTS` is safe to re-run):

| Object | Purpose |
|---|---|
| `idx_gnaf_pc`, `idx_cust_pc` | Index on `postcode` — supports the state/locality paths' coarser filtering. |
| `idx_gnaf_pcnum`, `idx_cust_pcnum` | Compound index on `(postcode, number_first)` — the index the hot postcode path actually relies on. |
| `idx_gnaf_label`, `idx_cust_label` | Index on `address_label` — backs the exact-label fast path. |
| `gnaf_locality_index` | Deduplicated `(locality_name, postcode, state)` — the small table the locality fallback's fuzzy suburb scan runs against, instead of scanning every address row. |
| `gnaf_match_cache` | `input_standardised → address_detail_pid` plus all six score components and a `cached_at` timestamp. |

`gnaf_init()` is always safe to re-run, including on a database built by an older
version — that's the entire point of the `IF NOT EXISTS` / `ADD COLUMN IF NOT
EXISTS` pattern throughout.

## The alias system

Three independent mechanisms populate non-core rows, all distinguished by the
`alias_type` column:

- **`street_only`** — derived locally by `gnaf_build_street_aliases()` from your own
  loaded `gnaf_addresses`: one row per unique `(street_name, street_type,
  street_suffix, locality_name, state, postcode)` combination, with `number_first`
  left `NULL`. PIDs are an MD5 of the key fields, so re-running without `overwrite =
  TRUE` is a no-op rather than a pile of duplicates.
- **`LOCALITY:SYN` / `STREET:SYN`** — loaded directly from G-NAF's own official alias
  tables by `gnaf_load_psv()`, when working from the raw "Standard" PSV product
  rather than the simplified Core CSV. These catch a *recognised alternative* suburb
  or street name that G-NAF itself records as a synonym — something the locally
  derived `street_only` aliases can't do, since they only ever drop the house number,
  never substitute a different name.
- **`alias_principal` / `principal_pid`** and **`primary_secondary` / `primary_pid`**
  — not alias *rows*, but columns on every row (core or alias) that capture G-NAF's
  own address relationships: whether a record is the canonical `PRINCIPAL` address or
  an `ALIAS` of one, and whether it's the main `PRIMARY` dwelling or a `SECONDARY`
  sub-dwelling/unit, with a pointer back to the related PID in each case.

`gnaf_match()`'s `alias_types` argument restricts which of these are eligible
candidates for a given call — `NA` selects core (non-alias) rows, and the default
`NULL` matches everything regardless of alias type.
